\section{Interpretation of QM}
\subsection{Supporting definitions}

To provide our interpretation of quantum mechanics we need to develop
a number of supporting definitions. As the reader familiar with
process algebraic systems can readily verify, these definitions make
\emph{essential} use of the reflective operations and as such identify
this calculus as uniquely suited to this particular task.

Among these operations we find a notion of \emph{multiplication} of
names that interacts well with a notion of \emph{tensor product} of
processes. Even more intriguingly, we find a notion of a \emph{dual}
to a process in the form of maps from processes to names. While
notions of composite names have been investigated in the process
algebraic literature, it is the fact that names reflect process
structure that enables the collection of duals to enjoy an algebraic
structure dual to the collection of processes (i.e. there are
operations available to duals that reflect the operations on
processes). Moreover, it is this structure that enables an effective
definition of inner product.

\subsubsection{Multiplication}
\begin{mathpar}
  \quotep{Q} \cdot \quotep{R} := \quotep{Q|R}
  \and \\
  \quotep{Q} \cdot P := P\{ \quotep{Q|R} / \quotep{R} : \quotep{R} \in \freenames{P} \}
\end{mathpar}

\paragraph{Discussion}
The first equation needs little explanation; the second says that each
free name of the process is replaced with the multiplication of that
name by the scalar. Multiplication of a scalar (name) by a state
(process) results in a process all the names of which have been `moved
over' by parallel composition with the process the scalar
quotes. There is a subtlety that the bound names have to be
manipulated so that multiplied names aren't accidentally
captured. There are many ways to achieve this.

\begin{remark}\label{rem:multiplication_identities}
  The reader is invited to verify that for all $x,y,z \in \QProc$ and $P \in \Proc$
  \begin{mathpar}
    x \cdot \quotep{0} \equiv x 
    \and
    x \cdot y \equiv y \cdot x
    \and
    x \cdot (y \cdot z) \equiv (x \cdot y) \cdot z
    \and \\
    \quotep{0} \cdot P \equiv P
    \and \\
    x \cdot (y \cdot P) \equiv (x \cdot y) \cdot P
    \and \\
    x \cdot (P|Q) \equiv (x \cdot P) | (x \cdot Q)
    \and \\    
  \end{mathpar}
\end{remark}

\subsubsection{Tensor product}

We define a tensor product on processes by structural induction.

\paragraph{Tensor of sums} First note that all summations, including
$\pzero$ and sequence, can be written $\Sigma_{i} x_{i}.A_{i} +
\Sigma_{j} x_{j}.C_{j}$, where we have grouped input-guarded processes
together and output-guarded processes together.

Thus, we can define the tensor product of two summations, $N_{1}\otimes N_{2}$, where

\begin{mathpar}
  N_{1} := \Sigma_{i} x_{i}.A_{i} + \Sigma_{j} x_{j}.C_{j}
  \and
  N_{2} := \Sigma_{i'} y_{i'}.B_{i'} + \Sigma_{j'} y_{j'}.D_{j'} 
\end{mathpar}

as follows.

\begin{mathpar}
  \Sigma_{i} x_{i}.A_{i} + \Sigma_{j} x_{j}.C_{j} \otimes \Sigma_{i'}
  y_{i'}.B_{i'} + \Sigma_{j'} y_{j'}.D_{j'} 
  \and \\
  := \; \Sigma_{i} \Sigma_{i'} \quotep{\stackrel{\vee}{x_{i}}| \stackrel{\vee}{y_{i'}}}.(A_{i}\otimes B_{i'}) \; | \; \Sigma_{i'} \Sigma_{i} \quotep{\stackrel{\vee}{y_{i'}}|\stackrel{\vee}{x_{i}}}.(B_{i'}\otimes A_{i})
  \and
  \;\; | \;\; \Sigma_{j} \Sigma_{j'} \quotep{\stackrel{\vee}{x_{j}}|\stackrel{\vee}{y_{j'}}}.(A_{j}\otimes B_{j'}) \; | \; \Sigma_{j'} \Sigma_{j} \quotep{\stackrel{\vee}{y_{j'}}|\stackrel{\vee}{x_{j}}}.(B_{j'}\otimes A_{j})
\end{mathpar}

\begin{remark}
  Do we need to $x^{L}$ and $y^{R}$ for this construction as well?
\end{remark}

\paragraph{Tensor of parallel compositions} Next, we distribute tensor
over par.

\begin{mathpar}
  P_{1}|P_{2} \otimes Q_{1}|Q_{2} := (P_{1} \otimes Q_{1}) | (P_{1}
  \otimes Q_{2}) | (P_{2} \otimes Q_{1}) | (P_{2} \otimes Q_{2})
\end{mathpar}

\paragraph{Tensor with dropped names} We treat tensor of a
process with a dropped name as parallel composition.

\begin{mathpar}
  P \otimes \dropn{x} := P | \dropn{x}
\end{mathpar}

\paragraph{Tensor of agents}

Finally, we need to define tensor on agents. Note that the definition
of tensor on summations only tensors inputs with inputs and outputs
with outputs. Thus, we only have to define the operation on
``homogeneous'' pairings.

\begin{mathpar}
  (\vec{x})P \otimes (\vec{y})Q
  \and \\
  := (x_{0}^{L}|y_{0}^{R},\ldots,x_{0}^{L}|y_{n}^{R},\ldots,x_{m}^{L}|y_{0}^{R},\ldots,x_{m}^{L}|y_{n}^R)(P\{ \vec{x}^{L}/\vec{x}\} \otimes Q \{ \vec{y}^{R}/\vec{y}\})
  \and \\
  \clift{\vec{P}} \otimes \clift{\vec{Q}}
  \and \\
  := \clift{P_{0}\otimes Q_{0},\ldots,P_{0}\otimes Q_{n},\ldots,P_{m}\otimes Q_{0},\ldots,P_{m}\otimes Q_{n}}
\end{mathpar}

\begin{remark}
  Observe that arities of tensored abstractions matches arities of
  tensored concretions if the original arities matched. Note also that
  the length of the arities corresponds to the increase in dimension
  we see in ordinary vector space tensor product.
\end{remark}

\begin{remark}
  Operationally, this definition distributes the tensor down to
  components ``linked'' by summation. Tensor over summation is
  intriguing in that it mixes names. Moreover, as a consequence of the
  way it mixes names we have the identities for all $x \in \QProc$ and
  $P,Q \in \Proc$

  \begin{mathpar}
    (x \cdot P) \otimes Q \equiv x \cdot (P \otimes Q) \equiv P \otimes (x \cdot Q)
    \and \\
    P \otimes \pzero \equiv P
  \end{mathpar}

  that the reader is invited to verify.
\end{remark}

\subsubsection{Annihilation}
\begin{mathpar}
  P^{\perp} := \{ Q : \forall R. P|Q \red^{*} R \Rightarrow R \red^{*} \pzero \}
  \and \\
  \annihilate{P} := \Sigma_{Q \in P^{\perp}} \quotep{Q}?(y).(\dropn{y}|Q) | \Sigma_{Q \in P^{\perp}} \quotep{Q}\clift{\Box}
\end{mathpar}

\paragraph{Discussion} The reader will note that $P^{\perp}$ is a
\emph{set} of processes, while $\annihilate{P}$ is a
\emph{context}. We call the set $P^{\perp}$ the \emph{annihilators} of
$P$. The parallel composition of a process in the annihilators of $P$
with $P$ will result in a process, the state space of which has all
paths eventually leading to $\pzero$. Execution may endure loops; but
under reasonable conditions of fairness (naturally guaranteed under
most notions of bisimulation) such a composite process cannot get
stuck in such a loop and will, eventually pop out and terminate.

The context $\annihilate{P}$ is ready and willing to ``take the
$P$ out of'' the process to which it is applied. It will effectively
transmit the code of the process to which it is applied to one of the
annihilators and run the process against it.

\begin{remark}
  Note that ${\annihilate{P}}^*$ is the abstraction corresponding to
  context $\annihilate{P}$. We will set $\dualize{P} := {\annihilate{P}}^*$.
\end{remark}

\subsubsection{Evaluation}
We fix $M$ a domain of fully abstract interpretation with an equality
coincident with bisimulation. We take $\meaningof{\cdot} : \Proc \to
M$ to be the map interpreting processes and $\nmeaningof{\cdot} : \M
\to Proc$ to be the map running the other way. Then we define

\begin{mathpar}
  \int P := \nmeaningof{\meaningof{P}}
\end{mathpar}

\paragraph{Discussion}
There are many fully abstract interpretations of Milner's
$\pi$-calculus. Any of them can be used as a basis for interpreting
the reflective calculus here. Equipped with such a domain it is
largely a matter of grinding through to check that the Yoneda
construction for the normalization-by-evaluation program can be
extended to this setting.

\begin{remark}
  The reader is invited to verify that $\int (\annihilate{P}[P]) = 0$,
  and equivalently $(\nu\; x)\int \dualize{P}\langle x \rangle |
  x\clift{P} = 0$.
\end{remark}

\subsection{Quantum mechanics}

\subsubsection{What is the quantum mechanical notion of continuation?}\label{sec:quantum_continuation}

Imagine the following experimental set-up. Alice, our intrepid quantum
investigator, prepares a state by performing some operation on some
initial state. Then she performs some measurement to obtain an
observation. Using the information of the observation, she selects a
new initial state, operation and measurement and repeats the steps
above. She iterates this procedure until she obtains some desired
observation. What is the expression of this procedure in the language
of quantum mechanics?

\begin{figure}[htp]\label{fig:iterated_experiment}
%  \fbox{
    \begin{lstlisting}[mathescape]
      $\mathcal{E} ::=$
      let S = $U \state{L}$ in (* prepare state*)
      let m = $\innerprod{M}{S}^2$ in (* take measurement *)
      match m with (* use m to decide next experiment *)
      v$_0$ -> $\mathcal{E}$
      | $\ldots$
      | v$_N$ -> $\mathcal{E}$
      | v$_{Exit}$ -> m (* return observation *)
    \end{lstlisting}
%    }
  \caption{Iterated experiment schematic}
\end{figure}

Figure \ref{fig:iterated_experiment} gives a schematic description of
such an iterated experimental procedure. The question is how do we
write down this iterated procedure without stepping outside the
language of quantum mechanics? Note that accounts of famous composite
quantum experiments, like the Stern-Gerlach experiment leave the
language of quantum mechanics to describe the iterated experiment.

We ask this question for many reasons, but one of them is to help set
up the exegesis of our interpretation. In our framework
\emph{everything} is a computation, both the quantum operations and
processes (classical or quantum) that invoke those operations. There
is no need to step out of the conceptual (and more pragmatically, the
computational) framework to describe these kinds of experiments. More
to the point, the framework we are proposing is -- like the hybrid
functional language employed in the schema -- \emph{compositional}:
experiments, computations are built out of experiments and
computations. This is of enormous pragmatic value if we are to build
and reason about systems of significant scale.

\begin{remark}
  It is also worth noting in this connection that this schema is the
  core of a wide range of recursive functions. Further, this
  connection to calculations of fixpoints makes it a close neighbor of
  search techniques like natural selection and the scientific
  method. This is a theme to which we will return, for quantum
  information seems very \emph{unlife-like} in it's uncloneable,
  undeleteable nature.
\end{remark}

Returning the matter of the computational interpretation, our
interpretation will take the form of a map, written
$\meaningof{-}(-)$, from expressions in Dirac notation to expressions
in our target reflective calculus. The map takes an \emph{ancillary}
argument, a channel along which to communicate results to subsequent
computations. This is how we communicate, for example, the results of
taking a measurement to a subsequent step in an experiment.

\subsubsection{Interpretation}

Table \ref{tbl:core_qm_op_defns} gives the core operational
correspondences. It is meant as an intuitive guide.

\begin{table}[htp]\label{tbl:core_qm_op_defns}
  \center{
    \fbox{
      \begin{tabular}{c|c}
        quantum mechanics & process calculus \\
        \hline
        scalar & $x := \quotep{P}$ \\
        state vector & $\state{P} := P$ \\
        dual & $\state{P}^{*} := \event{\annihilate{P}} := \quotep{\annihilate{P}}[-]$ \\
        matrix & $ \Sigma_{\alpha} \state{P_{\alpha}}x_{\alpha}\event{Q_{\alpha}}$ \\
        vector addition & $\state{P} + \state{Q} := \state{P | Q}$ \\
        tensor product & $\state{P} \otimes \state{Q} := \state{P \otimes Q}$ \\
        inner product & $\innerprod{P}{Q} := \quotep{\int \annihilate{P}[Q]}$ \\
      \end{tabular}
    }
  }
  \caption{QM - operational definitions}
\end{table}

\paragraph{Discussion}
The process algebraic view of a state is called, ironically, a
process, and that is what we map vectors to in our interpretation. It
has long been noted in the process algebraic community that names play
a role somewhat similar to scalars in a vector space. What is unique
about the reflective calculus, and makes it suitable for an
interpretation of this form is that with the structure of names
reflecting the structure of processes we can both make this similarity
in a precision instrument; and find a notion of \emph{dual} to a
state. 

If we posit names as scalars, then in perfect analogy with vector
spaces a dual is a map from processes to names. We actually have two
candidates for this interpretation: nominal contexts, $\quotep{M}$,
and their corresponding abstraction, $\quotep{M}^{*}$. The goal of
supporting a notion of continuation selects the latter of the two for
our interpretation.

Taking these as the basis of the interpretation together with the
algebraic identities required by the Dirac notation more or less fixes
the definitions of the rest of the operations. Among the interesting
particularities, the definition of inner product finds near perfect
mirroring of the Feyman interpretation.

\begin{mathpar}
  \inferrule* [lab=states] {} {\meaningof{\state{P}}(c) = c?(l,r).r\clift{P}}
  \and
  \inferrule* [lab=events] {} {\meaningof{\event{P}}(c) = (x)c?(l,r).l\clift{\dualize{P}\langle x \rangle} } 
  \and
  \inferrule* [lab=vector addition] {} {\meaningof{\state{P} + \state{Q}}(c) = \meaningof{\state{P | Q}}(c)}
  \and
  \inferrule* [lab=tensor product] {} {\meaningof{\state{P} \otimes \state{Q}}(c) = \meaningof{\state{P\otimes Q}}(c)}
  \and
  \inferrule* [lab=inner product] {} {\meaningof{\innerprod{P}{Q}}(c) = (\nu\; x)c\clift{\int \dualize{P}\langle x \rangle | x\clift{Q}}}
  \and
  \inferrule* [lab=matrix] {} {\meaningof{\fprmatrix{P}{x}{Q}}(c) = \\\\ (u)(\nu \; lr) c!(l,r).(l?(e). (\nu\; y)\meaningof{\innerprod{\dropn{e}}{Q}}(x) | x?(z).x?(a).c\clift{(\dualize{P}\sigma(z,a))\langle u \rangle}|y!(x) \\\\
    + r?(e). (\nu\; x)\meaningof{\innerprod{P}{\dropn{e}}}(x) | x?(z).x?(a).c\clift{Q\sigma(z,a)}|x!(x_{\alpha}))}  
  \and
  \inferrule* [lab=matrix application] {} {\meaningof{(\fprmatrix{P}{x}{Q})(\state{S})}(c) = (\nu\; c'u)\meaningof{\fprmatrix{P}{x}{Q}}(c')\langle u \rangle | \meaningof{\state{S}}(c') | c'?(a).c!(a)
  \\\\
  \meaningof{(\fprmatrix{P}{x}{Q})(\event{S})}(c) = (\nu\; c'u)\meaningof{\fprmatrix{P}{x}{Q}}(c')\langle u \rangle | \meaningof{\event{S}}(c') | c'?(a).c!(a)}
\end{mathpar}

where

\begin{mathpar}
  P\sigma(z,a) := P\{ z\cdot a\cdot r/r : r \in \freenames{P} \}
\end{mathpar}

\begin{remark}
  The reader is invited to verify that
  \begin{mathpar}
    \meaningof{\innerprod{P}{Q}}(c)
    \and \\
    \wbbisim 
    \and \\
    %\and
    (\nu\; x)(\nu c'lr)\meaningof{\event{P}}(c')\langle x \rangle \;|\; \meaningof{\state{Q}}(c')
    %\and
    |\; c'!(l,r).l?(p).r?(q).c\clift{\int \dropn{p} | x!(q)}
  \end{mathpar}
  This provides a (more) compositional definition of inner product. It
  also illustrates an important point of the computational
  interpretation. We have a notion of equivalence providing a crucial
  proof method: bisimulation.
\end{remark}

\begin{remark}
  Assuming $\int (\annihilate{P}[P]) = 0$, the reader is
  invited to verify that $(\fprmatrix{P}{x}{P})(\state{P}) = x \cdot \state{P}$.
\end{remark}

% \begin{remark}
%   The reader is invited to verify that $\innerprod{P}{Q}$ could
%   equally well have been written $\quotep{\int \stackrel{\vee}{x}}$
%   where $x = \event{\annihilate{P}}(Q)$.

%   One of the motivations for this remark is that there is another way
%   to factor these operations. We could package up evaluation in the dual:

%   \begin{mathpar}
%     \state{P}^{*} := \event{\int \annihilate{P}} := \quotep{\int \annihilate{P}}[-]
%   \end{mathpar}

%   and then have inner product defined by
  
%   \begin{mathpar}
%     \innerprod{M}{Q} := \event{M}(Q)
%   \end{mathpar}

%   where we use $M$ to label the dual to emphasize that it is a context.

%   Hopefully, experience with the calculations will provide guidance on
%   the best factoring.
% \end{remark}

\begin{remark}\label{rem:abstract_scalars}
  Assuming $\int (\annihilate{P}[P]) = 0$, the reader is
  invited to verify that $\forall P,Q. (\prmatrix{0}{Q})(\state{0}) =
  \state{0}$ and dually $(\prmatrix{P}{0})(\event{0}) = \event{0}$.
\end{remark}

\subsubsection{Interpreting continuations}

As promised, we can combine these interpretations with standard
semantics for conditionals and continuations to provide an
interpretation of the iterated experiment.

\begin{lstlisting}[mathescape]      
  $\ldb$ let S = $U \state{L}$ in 
  let m = $\innerprod{M}{S}^2$ in 
  match m with 
  v$_0$ -> $\mathcal{E}$
  | $\ldots$
  | v$_N$ -> $\mathcal{E}$
  | v$_{Exit}$ -> m $\rdb(c)$
  $=$
  $(\nu\; c')\meaningof{\innerprod{M}{U\state{L}}}(c')$
  $| c'?(m).m!(m) | \Sigma_{i=0}^{N}v_{i}?(m).\meaningof{\mathcal{E}} + v_{Exit}?(m).c!(m)$
\end{lstlisting}

\paragraph{A quick tally}
Already the interpretation is beginning to show signs of
promise. First of all, it is no more notationally cumbersome than the
notation used in QM calculations. Beyond syntax, we have a new proof
principle in hand and the ability to reason about more complex
experimental situations than is directly calculable in ordinary
quantum mechanics.

\subsubsection{Adjointness}

We need to give a definition of $(\cdot)^{\dagger}$ for matrices. The
obvious candidate definition is
\begin{mathpar}
\meaningof{(\fprmatrix{P}{x}{Q})^{\dagger}}(c)
= (\nu\; u)\meaningof{\fprmatrix{\dualize{Q}\langle u \rangle}{\overline{x}}{\dualize{P}\langle u \rangle}}(c) 
\end{mathpar}

% But, $(Q_{\alpha}^{\underline{\perp}})^{*}$ requires a name along
% which to communicate the process to achieve the context application.

\begin{remark}
  i'm a little worried that i don't (yet) have proper support for
  complex conjugacy. But, the observation above may give us a
  clue. According to Abramsky, it must be the case that the scalars
  are iso to the homset of the identity for the tensor -- which the
  observation above (\ref{rem:abstract_scalars}) characterizes. For
  now, we will simply bookmark the notion with $\overline{x}$.
\end{remark}

\subsubsection{Basis for a basis}
If processes label states and ``addition'' of states (a.k.a. vector
addition) is interpreted as parallel composition, what corresponds to
notions of linear independence and basis? Here, we recall that Yoshida
has developed a set of \emph{combinators} for an asynchronous verison
of Milner's $\pi$-calculus \cite{DBLP:conf/concur/Yoshida98}. These
are a finite set of processes such any process can be expressed as
parallel composition of these combinators together with liberal uses
of the new operator and replication. We can simply give a translation
of these into the present calculus and have reasonable expectation
that the property carries over. That is, that the resultant set allows
to express all processes via parallel composition. Note, however, that
there is no new operator or replication in this calculus. As a result,
we expect that the corresponding set is actually infinite. That is, we
expect that the space is actually infinite dimensional.

\begin{remark}
  The reader familiar with the lambda calculus may reasonably object:
  certainly, the collection $S$, $K$ and $I$ is a finite set of
  combinators \cite{Barendregt84}. Shouldn't we expect to see a finite
  set of combinators for an effectively equivalent system? i am very
  sympathetic to this critique and feel it warrants full attention. On
  the other hand, i also have in mind the following analogy. The
  natural numbers, as a monoid under addition, has exactly $1$
  generator, while the natural numbers, as a monoid under
  multiplication, has countably many generators (the primes). We
  observe that the application of the lambda calculus is much less
  resource sensitive than the parallel composition of the
  $\pi$-calculus. Could it be the case that we have an analogy of the
  form
  
  \begin{mathpar}
    m + n : MN :: m*n : M|N
  \end{mathpar}

  giving a similar blow up in the set of ``primes''?  This is such a
  wonderful thought that, even if it's not true, i think it's worth
  writing down.
\end{remark}
